cardinalR: Generating interesting high-dimensional data structures

### 2025-10-06 paper-cardinalR.Rmd
Updated the gen hole section

The package provides functionality for generating datasets with **spherical hole** (in 2D/3D) or, more generally, **hyperspherical hole** (in higher dimensions). These structures are valuable for evaluating how dimension reduction methods and clustering algorithms handle incomplete manifolds or missing regions of the data space. A hyperspherical hole introduces topological complexity: the structure remains continuous but contains excluded regions (voids) that algorithms must correctly represent in lower-dimensional embeddings.

The core function `gen_hole()` removes points from a dataset that fall within a user-specified hypersphere. Formally, given data points ($x \in \mathbb{R}^p$), a center ($a \in \mathbb{R}^p$), and radius ($r > 0$), only points satisfying

$$
||x - a||_2 > r
$$

are retained. The anchor point (a) can either be user-specified or default to the dataset mean, and radius (r) is controlled by the user, with safeguards to avoid trivial or degenerate cases. Because it operates generically on any dataset, spherical or hyperspherical holes can be embedded in a wide range of geometric structures.

Two specialized wrappers illustrate this idea. The function `gen_scurvehole()` generates an S-curve with a spherical hole by applying `gen_hole()` to the output of `gen_scurve()`. This structure has been used in prior diagnostic studies of NLDR methods, since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, `gen_unifcubehole()`, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.

>>>>

The package provides functionality for generating datasets with spherical hole (in $2\text{-}D$/$3\text{-}D$) or, more generally, hyperspherical hole (in higher dimensions). These structures are valuable for evaluating how dimension reduction methods and clustering algorithms handle incomplete manifolds or missing regions of the data space. A hyperspherical hole introduces topological complexity: the structure remains continuous but contains excluded regions (voids) that algorithms must correctly represent in lower-dimensional embeddings.

The core function `gen_hole(df, anchor, r)` removes points from a dataset that fall within a user-specified hypersphere. Formally, given data points ($x \in \mathbb{R}^p$), a center ($a \in \mathbb{R}^p$), and radius ($r > 0$), only points satisfying $||x - a||_2 > r$ are retained. The anchor point ($a$) can either be user-specified or default to the dataset mean, and radius ($r$) is controlled by the user, with safeguards to avoid trivial or degenerate cases. Because it operates generically on any dataset, spherical or hyperspherical holes can be embedded in a wide range of geometric structures.

Two specialized wrappers illustrate this idea. The function `gen_scurvehole(n, r_hole)` generates an S-curve with a spherical hole by applying `gen_hole()` to the output of `gen_scurve()`. This structure has been used in prior diagnostic studies of NLDR methods, since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, `gen_unifcubehole(n, p, r_hole)`, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.

### 2025-10-06 paper-cardinalR.tex
Updated the gen hole section

The package provides functionality for generating datasets with \textbf{spherical hole} (in 2D/3D) or, more generally, \textbf{hyperspherical hole} (in higher dimensions). These structures are valuable for evaluating how dimension reduction methods and clustering algorithms handle incomplete manifolds or missing regions of the data space. A hyperspherical hole introduces topological complexity: the structure remains continuous but contains excluded regions (voids) that algorithms must correctly represent in lower-dimensional embeddings.

The core function \texttt{gen\_hole()} removes points from a dataset that fall within a user-specified hypersphere. Formally, given data points (\(x \in \mathbb{R}^p\)), a center (\(a \in \mathbb{R}^p\)), and radius (\(r > 0\)), only points satisfying

\[
||x - a||_2 > r
\]

are retained. The anchor point (a) can either be user-specified or default to the dataset mean, and radius (r) is controlled by the user, with safeguards to avoid trivial or degenerate cases. Because it operates generically on any dataset, spherical or hyperspherical holes can be embedded in a wide range of geometric structures.

Two specialized wrappers illustrate this idea. The function \texttt{gen\_scurvehole()} generates an S-curve with a spherical hole by applying \texttt{gen\_hole()} to the output of \texttt{gen\_scurve()}. This structure has been used in prior diagnostic studies of NLDR methods, since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, \texttt{gen\_unifcubehole()}, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.

>>>>

The package provides functionality for generating datasets with spherical hole (in \(2\text{-}D\)/\(3\text{-}D\)) or, more generally, hyperspherical hole (in higher dimensions). These structures are valuable for evaluating how dimension reduction methods and clustering algorithms handle incomplete manifolds or missing regions of the data space. A hyperspherical hole introduces topological complexity: the structure remains continuous but contains excluded regions (voids) that algorithms must correctly represent in lower-dimensional embeddings.

The core function \texttt{gen\_hole(df,\ anchor,\ r)} removes points from a dataset that fall within a user-specified hypersphere. Formally, given data points (\(x \in \mathbb{R}^p\)), a center (\(a \in \mathbb{R}^p\)), and radius (\(r > 0\)), only points satisfying \(||x - a||_2 > r\) are retained. The anchor point (\(a\)) can either be user-specified or default to the dataset mean, and radius (\(r\)) is controlled by the user, with safeguards to avoid trivial or degenerate cases. Because it operates generically on any dataset, spherical or hyperspherical holes can be embedded in a wide range of geometric structures.

Two specialized wrappers illustrate this idea. The function \texttt{gen\_scurvehole(n,\ r\_hole)} generates an S-curve with a spherical hole by applying \texttt{gen\_hole()} to the output of \texttt{gen\_scurve()}. This structure has been used in prior diagnostic studies of NLDR methods, since it tests the ability of algorithms to capture non-linear manifolds that are not simply connected. The second wrapper, \texttt{gen\_unifcubehole(n,\ p,\ r\_hole)}, generates uniformly sampled cube data with a hyperspherical hole. By embedding a hyperspherical void inside a convex high-dimensional structure, this creates non-convex regions that challenge algorithms in terms of separability and neighborhood preservation.
